**OptiFit: a fast method for fitting amplicon sequences to existing OTUs**

### 2021-08-05 paper/manuscript.Rmd
final edits of first draft

**Other title options:**\
Optimal Taxonomic Resolution for Microbiome-Based Classification of Colorectal Cancer\
...

@@

Colorectal cancer is one of the most common cancers in men and women and a leading cause of cancer related deaths in the United States {}. Early detection and treatment are essential to increase survival rates, but for a variety of reasons including the invasiveness and high cost of screening (i.e. colonoscopy), many people do not comply with recommended screening guidelines {} prompting a need for low cost, non-invasive detection methods. A growing body of research points to the gut microbiome as a promising target for non-invasive screening to detect screen relevant neoplasias (neoplasms?) (SRNs) consisting of pre-cancerous polyps (i.e. advanced adenomas) and carcinomas. Efforts to realize the diagnostic potential of the gut microbiome in detecting SRNs have focused on machine learning (ML) methods using abundances from operational taxonomic unit (OTU) classifications based on amplicon sequencing of the 16S rRNA gene {}. However, whether this is the optimal taxonomic resolution for classifying SRNs from microbiome data is unknown. Additionally, recent work has pushed for the use of amplicon sequence variants (ASVs) to replace OTUs for marker-gene analysis because of the improved resolution with ASVs {}. However, whether the additional resolution provided by ASVs is useful for ML classification is unclear. Since ML classification relies on consistent differences between groups, its possible that the resolution at the ASV level is too individualized to accurately differentiate groups. Topçuoğlu *et al* [@topçuolu2020] recently demonstrated effective application of machine learning (ML) to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way (mikropml). This analysis utilizes the reproducible framework developed by Topçuoğlu *et al* to quantify which ML method and taxonomic level produce the best performing classifier for detecting SRNs from microbiome data.

@@

Utilizing publicly available 16S rRNA sequence data from patients with cancer and healthy controls, we generated abundance tables with Mothur {} annotated to Phylum, Class, Order, Family, Genus, OTU and ASV levels. Using the taxonomic abundance data, we quanitfied how accurately samples could be classified as "normal" or "SRN" (i.e. advanced adenoma or cancer) using five machine learning methods with the Mikropml R package (methods). Across the five machine learning methods tested, model performance tended to increase with taxonomic level usually peaking around genus/OTU level before dropping off slightly with ASVs (Supplemental Figure 1). Random forest (RF) was consistently the top performer at most taxonomic levels. Within the RF model, the highest AUROCs were observed for family (median AUROC: `r round(rf_med_auc["family"],digits=3)`), genus (median AUROC: `r round(rf_med_auc["genus"],digits=3)`), and OTU (median AUROC: `r round(rf_med_auc["otu"],digits=3)`) level data with no significant difference between the three (Figure 1). Performance with ASVs (median AUROC: `r round(rf_med_auc["asv"],digits=3)`) was significantly lower than OTU and genus (p \< 0.05) but not family level (p = 0.06). These results were consistent for ASVs generated with DADA2 {} (median AUROC: `r round(dada2_med_auc,digits=3)`). These results suggest that finer taxonomic resolution is not necessarily better for differentiating a persons cancer status from the microbiome.

@@

**is this paragraph useful???** *Of note, the ML pipeline includes a pre-processing step that occurs prior to training and classifying the ML models (methods). As part of this step, features are removed that will not provide useful information to build the model. For example, strongly correlated features provide the same information to build the model and thus can be collapsed. Additionally, features with zero or near-zero variance will not help the model differentiate groups and thus can be removed. Interestingly, despite starting with `r input %>% filter(level == "asv") %>% select(n_features) %>% pull()` features at the ASV level, only `r input %>% filter(level == "asv") %>% select(n_features_preproc) %>% pull()` (`r input %>% filter(level == "asv") %>% select(pct_kept) %>% pull()`%) remained after pre-processing. At the OTU level, `r input %>% filter(level == "otu") %>% select(n_features) %>% pull()` of the `r input %>% filter(level == "otu") %>% select(n_features_preproc) %>% pull()` features (`r input %>% filter(level == "otu") %>% select(pct_kept) %>% pull()`%) remained after preprocessing (Table 1).*

@@

These results demonstrate that consideration of the appropriate taxonomic resolution for utilizing the microbiome as a predictive tool. In general, we found that finer resolution (e.g. ASVs) does not add additional sensitivity to model predicition and at the ASV level actually impedes model performance due to the sparsity of shared taxa. Additionally, utilizing genus or family level data could allow for merging data generated from different 16S regions or sequencing platforms.

-   Overall, the fine resolution of ASVs is unnecessary for ML classification with microbiome data since the vast majority of the ASVs are removed during preprocessing and the model performance is diminshed compared to other taxonomic levels.
-   since family/genus level is just as good as otu could we use data from different 16S regions?

@@

```{r feats}

>>>>

@@

Colorectal cancer is one of the most common cancers in men and women and a leading cause of cancer related deaths in the United States [@siegel2020]. Early detection and treatment are essential to increase survival rates, but for a variety of reasons including the invasiveness and high cost of screening (i.e. colonoscopy), many people do not comply with recommended screening guidelines [@garcía2011a] prompting a need for low cost, non-invasive detection methods. A growing body of research points to the gut microbiome as a promising target for non-invasive screening to detect screen relevant neoplasias (neoplasms?) (SRNs) consisting of pre-cancerous polyps (i.e. advanced adenomas) and carcinomas. Efforts to realize the diagnostic potential of the gut microbiome in detecting SRNs have focused on machine learning (ML) methods using abundances from operational taxonomic unit (OTU) classifications based on amplicon sequencing of the 16S rRNA gene {}. However, whether this is the optimal taxonomic resolution for classifying SRNs from microbiome data is unknown. Additionally, recent work has pushed for the use of amplicon sequence variants (ASVs) to replace OTUs for marker-gene analysis because of the improved resolution with ASVs [@callahan2017]. However, whether the additional resolution provided by ASVs is useful for ML classification is unclear {pat preprint?}. Since ML classification relies on consistent differences between groups, its possible that the resolution at the ASV level is too individualized to accurately differentiate groups. Topçuoğlu *et al* [@topçuolu2020] recently demonstrated effective application of machine learning (ML) to microbiome based classification problems and developed a framework for applying ML practices in a more reproducible way (mikropml). This analysis utilizes the reproducible framework developed by Topçuoğlu *et al* to quantify which ML method and taxonomic level produce the best performing classifier for detecting SRNs from microbiome data.

@@

Utilizing publicly available 16S rRNA sequence data from stool of patients with SRNs and healthy controls, we generated abundance tables with Mothur {} annotated to Phylum, Class, Order, Family, Genus, OTU and ASV levels. Using the taxonomic abundance data, we quantified how accurately samples could be classified as "normal" or "SRN" (i.e. advanced adenoma or carcinoma) using five machine learning methods with the Mikropml R package (methods). Across the five machine learning methods tested, model performance tended to increase with taxonomic level usually peaking around genus/OTU level before dropping off slightly with ASVs (Supplemental Figure 1). Random forest (RF) was consistently the top performer at most taxonomic levels. Within the RF model, the highest AUROCs were observed for family (median AUROC: `r round(rf_med_auc["family"],digits=3)`), genus (median AUROC: `r round(rf_med_auc["genus"],digits=3)`), and OTU (median AUROC: `r round(rf_med_auc["otu"],digits=3)`) level data with no significant difference between the three (Figure 1). Performance with ASVs (median AUROC: `r round(rf_med_auc["asv"],digits=3)`) was significantly lower than OTU and genus (p \< 0.05) but not family level (p = 0.06). These results were consistent for ASVs generated with DADA2 {} (median AUROC: `r round(dada2_med_auc,digits=3)`). These results suggest that finer taxonomic resolution is not necessarily better for differentiating a persons cancer status from the microbiome.

@@

Of note, the ML pipeline includes a pre-processing step that occurs prior to training and classifying the ML models (methods). As part of this step, features are removed that will not provide useful information to build the model. For example, strongly correlated features provide the same information to build the model and thus can be collapsed. Additionally, features with zero or near-zero variance will not help the model differentiate groups and thus can be removed. Interestingly, despite starting with `r input %>% filter(level == "asv") %>% select(n_features) %>% pull()` features at the ASV level, only `r input %>% filter(level == "asv") %>% select(n_features_preproc) %>% pull()` (`r input %>% filter(level == "asv") %>% select(pct_kept) %>% pull()`%) remained after pre-processing. At the OTU level, `r input %>% filter(level == "otu") %>% select(n_features) %>% pull()` of the `r input %>% filter(level == "otu") %>% select(n_features_preproc) %>% pull()` features (`r input %>% filter(level == "otu") %>% select(pct_kept) %>% pull()`%) remained after preprocessing (Table 1).

@@

These results demonstrate that consideration of the appropriate taxonomic resolution for utilizing the microbiome as a predictive tool. In general, we found that finer resolution (e.g. ASVs) does not add additional sensitivity to model predicition and at the ASV level actually impedes model performance due to the sparsity of shared taxa.

-   Utilizing genus or family level data could allow for merging data generated from different 16S regions or sequencing platforms. (is this true?)

While predictive models utilizing microbiome data alone are not that great, microbiome in combination with FIT results are better than either alone {}. This analysis wanted to quantify difference between

@@

```{r feature_counts}

### 2021-08-05 paper/manuscript.tex
final edits of first draft

\textbf{Other title options:}\\
Optimal Taxonomic Resolution for Microbiome-Based Classification of
Colorectal Cancer\\
\ldots{}

@@

Colorectal cancer is one of the most common cancers in men and women and
a leading cause of cancer related deaths in the United States \{\}.
Early detection and treatment are essential to increase survival rates,
but for a variety of reasons including the invasiveness and high cost of
screening (i.e.~colonoscopy), many people do not comply with recommended
screening guidelines \{\} prompting a need for low cost, non-invasive
detection methods. A growing body of research points to the gut
microbiome as a promising target for non-invasive screening to detect
screen relevant neoplasias (neoplasms?) (SRNs) consisting of
pre-cancerous polyps (i.e.~advanced adenomas) and carcinomas. Efforts to
realize the diagnostic potential of the gut microbiome in detecting SRNs
have focused on machine learning (ML) methods using abundances from
operational taxonomic unit (OTU) classifications based on amplicon
sequencing of the 16S rRNA gene \{\}. However, whether this is the
optimal taxonomic resolution for classifying SRNs from microbiome data
is unknown. Additionally, recent work has pushed for the use of amplicon
sequence variants (ASVs) to replace OTUs for marker-gene analysis
because of the improved resolution with ASVs \{\}. However, whether the
additional resolution provided by ASVs is useful for ML classification
is unclear. Since ML classification relies on consistent differences
between groups, its possible that the resolution at the ASV level is too
individualized to accurately differentiate groups. Topçuoğlu \emph{et
al} (\protect\hyperlink{ref-topuxe7uolu2020}{Topçuoğlu et al. 2020})
recently demonstrated effective application of machine learning (ML) to
microbiome based classification problems and developed a framework for
applying ML practices in a more reproducible way (mikropml). This
analysis utilizes the reproducible framework developed by Topçuoğlu
\emph{et al} to quantify which ML method and taxonomic level produce the
best performing classifier for detecting SRNs from microbiome data.

@@

Utilizing publicly available 16S rRNA sequence data from patients with
cancer and healthy controls, we generated abundance tables with Mothur
\{\} annotated to Phylum, Class, Order, Family, Genus, OTU and ASV
levels. Using the taxonomic abundance data, we quanitfied how accurately
samples could be classified as ``normal'' or ``SRN'' (i.e.~advanced
adenoma or cancer) using five machine learning methods with the Mikropml
R package (methods). Across the five machine learning methods tested,
model performance tended to increase with taxonomic level usually
peaking around genus/OTU level before dropping off slightly with ASVs
(Supplemental Figure 1). Random forest (RF) was consistently the top
performer at most taxonomic levels. Within the RF model, the highest
AUROCs were observed for family (median AUROC: 0.687), genus (median
AUROC: 0.686), and OTU (median AUROC: 0.698) level data with no
significant difference between the three (Figure 1). Performance with
ASVs (median AUROC: 0.676) was significantly lower than OTU and genus (p
\textless{} 0.05) but not family level (p = 0.06). These results were
consistent for ASVs generated with DADA2 \{\} (median AUROC: 0.66).
These results suggest that finer taxonomic resolution is not necessarily
better for differentiating a persons cancer status from the microbiome.

@@

\textbf{is this paragraph useful???} \emph{Of note, the ML pipeline
includes a pre-processing step that occurs prior to training and
classifying the ML models (methods). As part of this step, features are
removed that will not provide useful information to build the model. For
example, strongly correlated features provide the same information to
build the model and thus can be collapsed. Additionally, features with
zero or near-zero variance will not help the model differentiate groups
and thus can be removed. Interestingly, despite starting with 104106
features at the ASV level, only 478 (0.5\%) remained after
pre-processing. At the OTU level, 20079 of the 705 features (3.5\%)
remained after preprocessing (Table 1).}

@@

These results demonstrate that consideration of the appropriate
taxonomic resolution for utilizing the microbiome as a predictive tool.
In general, we found that finer resolution (e.g.~ASVs) does not add
additional sensitivity to model predicition and at the ASV level
actually impedes model performance due to the sparsity of shared taxa.
Additionally, utilizing genus or family level data could allow for
merging data generated from different 16S regions or sequencing
platforms.

\begin{itemize}
\tightlist
\item
  Overall, the fine resolution of ASVs is unnecessary for ML
  classification with microbiome data since the vast majority of the
  ASVs are removed during preprocessing and the model performance is
  diminshed compared to other taxonomic levels.
\item
  since family/genus level is just as good as otu could we use data from
  different 16S regions?
\end{itemize}

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-topuxe7uolu2020}{}%
Topçuoğlu, Begüm D., Nicholas A. Lesniak, Mack T. Ruffin, Jenna Wiens,
and Patrick D. Schloss. 2020. {``A Framework for Effective Application
of Machine Learning to Microbiome-Based Classification Problems.''}
\emph{mBio} 11 (3). \url{https://doi.org/10.1128/mBio.00434-20}.

>>>>

@@

Colorectal cancer is one of the most common cancers in men and women and
a leading cause of cancer related deaths in the United States
(\protect\hyperlink{ref-siegel2020}{Siegel et al. 2020}). Early
detection and treatment are essential to increase survival rates, but
for a variety of reasons including the invasiveness and high cost of
screening (i.e.~colonoscopy), many people do not comply with recommended
screening guidelines (\protect\hyperlink{ref-garcuxeda2011a}{García and
Z 2011}) prompting a need for low cost, non-invasive detection methods.
A growing body of research points to the gut microbiome as a promising
target for non-invasive screening to detect screen relevant neoplasias
(neoplasms?) (SRNs) consisting of pre-cancerous polyps (i.e.~advanced
adenomas) and carcinomas. Efforts to realize the diagnostic potential of
the gut microbiome in detecting SRNs have focused on machine learning
(ML) methods using abundances from operational taxonomic unit (OTU)
classifications based on amplicon sequencing of the 16S rRNA gene \{\}.
However, whether this is the optimal taxonomic resolution for
classifying SRNs from microbiome data is unknown. Additionally, recent
work has pushed for the use of amplicon sequence variants (ASVs) to
replace OTUs for marker-gene analysis because of the improved resolution
with ASVs (\protect\hyperlink{ref-callahan2017}{Callahan, McMurdie, and
Holmes 2017}). However, whether the additional resolution provided by
ASVs is useful for ML classification is unclear \{pat preprint?\}. Since
ML classification relies on consistent differences between groups, its
possible that the resolution at the ASV level is too individualized to
accurately differentiate groups. Topçuoğlu \emph{et al}
(\protect\hyperlink{ref-topuxe7uolu2020}{Topçuoğlu et al. 2020})
recently demonstrated effective application of machine learning (ML) to
microbiome based classification problems and developed a framework for
applying ML practices in a more reproducible way (mikropml). This
analysis utilizes the reproducible framework developed by Topçuoğlu
\emph{et al} to quantify which ML method and taxonomic level produce the
best performing classifier for detecting SRNs from microbiome data.

@@

Utilizing publicly available 16S rRNA sequence data from stool of
patients with SRNs and healthy controls, we generated abundance tables
with Mothur \{\} annotated to Phylum, Class, Order, Family, Genus, OTU
and ASV levels. Using the taxonomic abundance data, we quantified how
accurately samples could be classified as ``normal'' or ``SRN''
(i.e.~advanced adenoma or carcinoma) using five machine learning methods
with the Mikropml R package (methods). Across the five machine learning
methods tested, model performance tended to increase with taxonomic
level usually peaking around genus/OTU level before dropping off
slightly with ASVs (Supplemental Figure 1). Random forest (RF) was
consistently the top performer at most taxonomic levels. Within the RF
model, the highest AUROCs were observed for family (median AUROC:
0.687), genus (median AUROC: 0.686), and OTU (median AUROC: 0.698) level
data with no significant difference between the three (Figure 1).
Performance with ASVs (median AUROC: 0.676) was significantly lower than
OTU and genus (p \textless{} 0.05) but not family level (p = 0.06).
These results were consistent for ASVs generated with DADA2 \{\} (median
AUROC: 0.66). These results suggest that finer taxonomic resolution is
not necessarily better for differentiating a persons cancer status from
the microbiome.

@@

Of note, the ML pipeline includes a pre-processing step that occurs
prior to training and classifying the ML models (methods). As part of
this step, features are removed that will not provide useful information
to build the model. For example, strongly correlated features provide
the same information to build the model and thus can be collapsed.
Additionally, features with zero or near-zero variance will not help the
model differentiate groups and thus can be removed. Interestingly,
despite starting with 104106 features at the ASV level, only 478 (0.5\%)
remained after pre-processing. At the OTU level, 20079 of the 705
features (3.5\%) remained after preprocessing (Table 1).

@@

These results demonstrate that consideration of the appropriate
taxonomic resolution for utilizing the microbiome as a predictive tool.
In general, we found that finer resolution (e.g.~ASVs) does not add
additional sensitivity to model predicition and at the ASV level
actually impedes model performance due to the sparsity of shared taxa.

\begin{itemize}
\tightlist
\item
  Utilizing genus or family level data could allow for merging data
  generated from different 16S regions or sequencing platforms. (is this
  true?)
\end{itemize}

While predictive models utilizing microbiome data alone are not that
great, microbiome in combination with FIT results are better than either
alone \{\}. This analysis wanted to quantify difference between

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-callahan2017}{}%
Callahan, Benjamin J., Paul J. McMurdie, and Susan P. Holmes. 2017.
{``Exact Sequence Variants Should Replace Operational Taxonomic Units in
Marker-Gene Data Analysis.''} \emph{The ISME Journal} 11 (12): 2639--43.
\url{https://doi.org/10.1038/ismej.2017.119}.

\leavevmode\hypertarget{ref-garcuxeda2011a}{}%
García, Gimeno, and Antonio Z. 2011. {``Factors Influencing Colorectal
Cancer Screening Participation.''} \emph{Gastroenterology Research and
Practice} 2012 (December): e483417.
\url{https://doi.org/10.1155/2012/483417}.

\leavevmode\hypertarget{ref-siegel2020}{}%
Siegel, Rebecca L., Kimberly D. Miller, Ann Goding Sauer, Stacey A.
Fedewa, Lynn F. Butterly, Joseph C. Anderson, Andrea Cercek, Robert A.
Smith, and Ahmedin Jemal. 2020. {``Colorectal Cancer Statistics,
2020.''} \emph{CA: A Cancer Journal for Clinicians} 70 (3): 145--64.
https://doi.org/\url{https://doi.org/10.3322/caac.21601}.

\leavevmode\hypertarget{ref-topuxe7uolu2020}{}%
Topçuoğlu, Begüm D., Nicholas A. Lesniak, Mack T. Ruffin, Jenna Wiens,
and Patrick D. Schloss. 2020. {``A Framework for Effective Application
of Machine Learning to Microbiome-Based Classification Problems.''}
\emph{mBio} 11 (3). \url{https://doi.org/10.1128/mBio.00434-20}.
